The Rockerverse: Packages and Applications for Containerisation with R

### 2020-08-17 rockerverse.tex
update tex

First, different packages use containers for the \textbf{deployment of R
and \pkg{Shiny} apps}. \pkg{Shiny} is a popular package for creating
interactive online dashboards with R, and it enables users with very
diverse backgrounds to create stable and user-friendly web applications
\citep{cran_shiny}. \emph{ShinyProxy} (\url{https://www.shinyproxy.io/})
is an open-source tool to deploy Shiny apps in an enterprise context,
where it features single sign-on, but it can also be used in scientific
use cases
\citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
ShinyProxy uses Docker containers to isolate user sessions and to
achieve scalability for multi-user scenarios with multiple apps.
ShinyProxy itself is written in Java to accommodate corporate
requirements and may itself run in a container for stability and
availability. The tool is built on ContainerProxy
(\url{https://www.containerproxy.io/}), which provides similar features
for executing long-running R jobs or interactive R sessions. The started
containers can run on a regular Docker host but also in clusters.
Continuous integration and deployment (CI/CD) for Shiny applications
using Shinyproxy can be achieved, e.g., via
\href{https://docs.gitlab.com/ee/ci/pipelines.html}{GitLab pipelines}. A
pipeline can include building and checking R packages and Shiny apps.
After the code has passed the checks, Docker images are built and pushed
to GitLab's container registry. The pipeline finishes with triggering a
webhook on the server, where the deployment script is executed. The
script updates configurations, pulls the new Docker images, and restarts
the services.

@@

\citet{gentleman_statistical_2007} coined the term \textbf{Research
Compendium} for a dynamic document together with supporting data and
code. They used the R package system \citep{core_writing_1999} for the
functional prototype all the way to structuring, validating, and
distributing research compendia. This concept has been taken up and
extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.},
not in the least by applying containerisation and other methods for
managing computing environments---see Section~\nameref{envs}. Containers
give the researcher an isolated environment to assemble these research
pipelines with specific versions of software to minimize problems with
breaking changes and make workflows easier to share
\citep[cf.][]{boettiger_introduction_2015,marwick_packaging_2018}.
Research workflows in containers are safe from contamination from other
activities that occur on the researcher's computer, for example the
installation of the newest version of packages for teaching
demonstrations or specific versions for evaluation of others' works.
Given the users in this scenario, i.e., often academics with limited
formal software development training, templates and assistance with
containers around research compendia is essential. In many fields, we
see that a typical unit of research for a container is a research report
or journal article, where the container holds the compendium, or
self-contained set of data (or connections to data elsewhere) and code
files needed to fully reproduce the article
\citep{marwick_packaging_2018}. The package \pkg{rrtools}
(\url{https://github.com/benmarwick/rrtools}) provides a template and
convenience functions to apply good practices for research compendia,
including a starter \texttt{Dockerfile}. Images of compendium containers
can be hosted on services such as Docker~Hub for convenient sharing
among collaborators and others. Similarly, packages such as
\pkg{containerit} and \pkg{dockerfiler} can be used to manage the
\texttt{Dockerfile} to be archived with a compendium on a data
repository (e.g.~\href{https://zenodo.org/}{Zenodo},
\href{https://dataverse.org/}{Dataverse},
\href{https://figshare.com/}{Figshare}, \href{https://osf.io/}{OSF}). A
typical compendium's \texttt{Dockerfile} will pull a rocker image fixed
to a specific version of R, and install R packages from the MRAN
repository to ensure the package versions are tied to a specific date,
rather than the most recent version. A more extreme case is the
\emph{dynverse} project
\citep{saelens_comparisonsinglecelltrajectory_2019}, which packages over
50 computational methods with different environments (R, Python, C++,
etc.) in Docker images, which can be executed from R. \emph{dynverse}
uses a CI platform (see~\nameref{ci}) to build Rocker-derived images,
test them, and, if the tests succeed, publish them on Docker~Hub.

@@

The ordering of authors following DN and DE is alphabetical. DN
conceived the article idea,
\href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team},
wrote sections not mentioned below, and revised all sections. DE wrote
the introduction and the section about containerisation and the Rocker
Project, and reviewed all sections. DB wrote the section on
\pkg{outsider}. GD contributed the CARD.com use case. RC contributed to
the section on interfaces for Docker in R (\emph{dynverse} and
\texttt{dynwrap}). DC contributed content on Gigantum. ME contributed to
the section on processing and deployment to cloud services. CF wrote
paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and
\pkg{r-db}. EH contributed content on \pkg{dockyard}. LK contributed
content on \pkg{dockr}. SL contributed content on RStudio's usage of
Docker. BM wrote the section on research compendia and made the project
Binder-ready. HN \& JN co-wrote the section on the T-Mobile use case. KR
wrote the section about \pkg{holepunch}. NR wrote paragraphs about
shared work environments and GPUs. LS \& NT wrote the section on
Bioconductor. PS wrote the paragraphs about GitLab CI/CD pipelines with
Shinyproxy and OpenFaaS templates. TS \& JW wrote the section on
CyVerse. CvP wrote the section on the usage of Docker containers in
Dodona. CW wrote the sections on Whole Tale and contributed content
about publication reproducibility audits. NX contributed content on
\pkg{liftr}. All authors approved the final version. This articles was
collaboratively written at
\href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The
\href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}
and
\href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues}
provide details on the respective contributions.

>>>>

First, different packages use containers for the \textbf{deployment of R
and \pkg{Shiny} apps}. \pkg{Shiny} is a popular package for creating
interactive online dashboards with R, and it enables users with very
diverse backgrounds to create stable and user-friendly web applications
\citep{cran_shiny}. \emph{ShinyProxy} (\url{https://www.shinyproxy.io/})
is an open-source tool to deploy Shiny apps in an enterprise context,
where it features single sign-on, but it can also be used in scientific
use cases
\citep[e.g., ][]{savini_epiexplorer_2019,glouzon_structurexplor_2017}.
ShinyProxy uses Docker containers to isolate user sessions and to
achieve scalability for multi-user scenarios with multiple apps.
ShinyProxy itself is written in Java to accommodate corporate
requirements and may itself run in a container for stability and
availability. The tool is built on ContainerProxy
(\url{https://www.containerproxy.io/}), which provides similar features
for executing long-running R jobs or interactive R sessions. The started
containers can run on a regular Docker host but also in clusters.
Continuous integration and deployment (CI/CD) for Shiny applications
using Shinyproxy can be achieved, e.g., via GitLab pipelines or with a
combination of GitHub and Docker Hub. A pipeline can include building
and checking R packages and Shiny apps. After the code has passed the
checks, Docker images are built and pushed to the container registry.
The pipeline finishes with triggering a
\href{https://github.com/analythium/shinyproxy-1-click/blob/master/digitalocean/setup.md\#setting-up-webhook}{webhook}
on the server, where the deployment script is executed. The script can
update configurations or pull the new Docker images. There is a
\href{https://marketplace.digitalocean.com/apps/shinyproxy}{ShinyProxy
1-Click App} in the DigitalOcean marketplace that is set up with these
webhooks. The
\href{https://github.com/analythium/shinyproxy-1-click/blob/master/digitalocean/secure.md}{documentation}
explains how to set up HTTPS with ShinyProxy and webhooks.

@@

\citet{gentleman_statistical_2007} coined the term \textbf{Research
Compendium} for a dynamic document together with supporting data and
code. They used the R package system \citep{core_writing_1999} for the
functional prototype all the way to structuring, validating, and
distributing research compendia. This concept has been taken up and
extended\footnote{See full literature list at \href{https://research-compendium.science/}{https://research-compendium.science/}.},
not in the least by applying containerisation and other methods for
managing computing environments---see Section~\nameref{envs}. Containers
give the researcher an isolated environment to assemble these research
pipelines with specific versions of software to minimize problems with
breaking changes and make workflows easier to share
\citep[cf.][]{boettiger_introduction_2015,marwick_packaging_2018}.
Research workflows in containers are safe from contamination from other
activities that occur on the researcher's computer, for example the
installation of the newest version of packages for teaching
demonstrations or specific versions for evaluation of others' works.
Given the users in this scenario, i.e., often academics with limited
formal software development training, templates and assistance with
containers around research compendia is essential. In many fields, we
see that a typical unit of research for a container is a research report
or journal article, where the container holds the compendium, or
self-contained set of data (or connections to data elsewhere) and code
files needed to fully reproduce the article
\citep{marwick_packaging_2018}. The package \pkg{rrtools}
(\url{https://github.com/benmarwick/rrtools}) provides a template and
convenience functions to apply good practices for research compendia,
including a starter \texttt{Dockerfile}. Images of compendium containers
can be hosted on services such as Docker~Hub for convenient sharing
among collaborators and others. Similarly, packages such as
\pkg{containerit} and \pkg{dockerfiler} can be used to manage the
\texttt{Dockerfile} to be archived with a compendium on a data
repository (e.g. \href{https://zenodo.org/}{Zenodo},
\href{https://dataverse.org/}{Dataverse},
\href{https://figshare.com/}{Figshare}, \href{https://osf.io/}{OSF}). A
typical compendium's \texttt{Dockerfile} will pull a rocker image fixed
to a specific version of R, and install R packages from the MRAN
repository to ensure the package versions are tied to a specific date,
rather than the most recent version. A more extreme case is the
\emph{dynverse} project
\citep{saelens_comparisonsinglecelltrajectory_2019}, which packages over
50 computational methods with different environments (R, Python, C++,
etc.) in Docker images, which can be executed from R. \emph{dynverse}
uses a CI platform (see~\nameref{ci}) to build Rocker-derived images,
test them, and, if the tests succeed, publish them on Docker~Hub.

@@

The ordering of authors following DN and DE is alphabetical. DN
conceived the article idea,
\href{https://github.com/nuest/rockerverse-paper/issues/3}{initialised the formation of the writing team},
wrote sections not mentioned below, and revised all sections. DE wrote
the introduction and the section about containerisation and the Rocker
Project, and reviewed all sections. DB wrote the section on
\pkg{outsider}. GD contributed the CARD.com use case. RC contributed to
the section on interfaces for Docker in R (\emph{dynverse} and
\texttt{dynwrap}). DC contributed content on Gigantum. ME contributed to
the section on processing and deployment to cloud services. CF wrote
paragraphs about \pkg{r-online}, \pkg{dockerfiler}, \pkg{r-ci} and
\pkg{r-db}. EH contributed content on \pkg{dockyard}. LK contributed
content on \pkg{dockr}. SL contributed content on RStudio's usage of
Docker. BM wrote the section on research compendia and made the project
Binder-ready. HN \& JN co-wrote the section on the T-Mobile use case. KR
wrote the section about \pkg{holepunch}. NR wrote paragraphs about
shared work environments and GPUs. LS \& NT wrote the section on
Bioconductor. PS wrote the paragraphs about CI/CD pipelines with
Shinyproxy 1-Click app and OpenFaaS templates. TS \& JW wrote the
section on CyVerse. CvP wrote the section on the usage of Docker
containers in Dodona. CW wrote the sections on Whole Tale and
contributed content about publication reproducibility audits. NX
contributed content on \pkg{liftr}. All authors approved the final
version. This articles was collaboratively written at
\href{https://github.com/nuest/rockerverse-paper/}{https://github.com/nuest/rockerverse-paper/}.
The
\href{https://github.com/nuest/rockerverse-paper/graphs/contributors}{contributors page}
and
\href{https://github.com/nuest/rockerverse-paper/issues/}{discussion issues}
provide details on the respective contributions.
